---
title: "Figures for paper"
author: "Michael Bedward (mbedward@uow.edu.au)"
date: "17/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr, warn.conflicts = FALSE)
library(stringr)

library(rjags)
library(R2jags)

library(ggplot2)
library(ggrepel)


# Random number seed
set.seed(42)

# Set this to TRUE to force re-fitting of models if
# cached objects exist. Set to FALSE to use cached outputs.
REFIT_MODELS <- FALSE

# Default theme for graphs
theme_set( theme_bw() + theme(panel.grid = element_blank()))

# Load common functions
source(file = here("scripts", "common_functions.R"))


# Function to generate initial values for Z: matrix of probit values.
# Takes the matrix of sites x species occurrences as an argument.
#
fn_Zinit <- function(y) {
  nc <- ncol(y)
  nr <- nrow(y)
  
  Tau <- rWishart(1, nc + 1, diag(nc))[, , 1]
  Sigma <- solve(Tau)

  Z <- abs(t(mvtnorm::rmvnorm(nr, rep(0, nc), Sigma)))
  Z <- ifelse(as.matrix(y), Z, -1 * Z)
  
  Z
}

```


## Summary

This markdown document contains code to produce graphs for figures in the manuscript. It relies on data and fitted model objects that were (or should have been) saved to file from code in other documents.


## Data

Lookup tables for study sites and plots, and tree size classes (from the RRGcommon package used for other analyses).

```{r}

SITES <- load_from(here("data", "SITES.RData"))

SIZES <- load_from(here("data", "SIZE_CLASSES.RData"))

```


Response data is a matrix of species occurrences within 9ha experimental plots. The data for each plot are formed by pooling records from the three floristics quadrats located within the plot.

```{r}

DAT.occurrence.long <- load_from( here("data", "site_species_boral.RData") ) %>%
  
  mutate(siteplot = paste0(site, "-", plot9ha)) %>%
  
  distinct(site, plot9ha, siteplot, workingcode)

```


Coding, name and habitat information for species.

```{r}

SPECIES.info <- load_from( here("data", "species_info_boral.RData") )

```


Summary of species prevalence in sites, plots and quadrats.

```{r}

SPECIES.summary <- load_from( here("data", "species_summary_boral.RData") )

```


Flood history plot groupings.

```{r}

DAT.floodgroup <- load_from( here("data", "flood_group_boral.RData") )

```


### Identify a subset of species for modelling

Of the 200 species identified as candidates for modelling, 84 were only recorded in 1 or 2 experimental plots (most of the latter being in a single site). 

The threshold number of occurrences set to include a species for modelling is hard to define, but given the complexity of the models to be fitted, a minimum of 5 plot occurrences is a reasonable starting point.

We also remove Eucalpytus camaldulensis (present in all sites) from the data set.

```{r}

SpeciesSubset <- SPECIES.summary %>%
  dplyr::filter(nplots >= 5) %>%
  
  left_join(SPECIES.info, by = c("workingcode", "workingname")) %>%
  
  # remove red gums
  dplyr::filter(workingname != "Eucalyptus camaldulensis")

cat("Subset with", nrow(SpeciesSubset), "taxa", 
    "of which", sum(SpeciesSubset$weed), "are exotic \n")

```

### Species occurrence data matrix for modelling

```{r}

DAT.occurrence.subset <- DAT.occurrence.long %>%
  dplyr::filter(workingcode %in% SpeciesSubset$workingcode)

# Format response data as a matrix for boral
M.y <- DAT.occurrence.subset %>%
  mutate(present = 1) %>%
  
  tidyr::pivot_wider(names_from = workingcode, 
                     values_from = present, values_fill = 0) %>%

  # Order should be fine but just in case...
  arrange(site, plot9ha)


SiteLookup <- M.y %>%
  dplyr::select(site, plot = plot9ha, siteplot) %>%
  left_join(
    SITES %>% dplyr::select(site, plot, site.quality),
    by = c("site", "plot")
  ) %>%
  
  left_join(DAT.floodgroup, by = c("site", "plot", "siteplot"))


M.y <- as.matrix(M.y[ , -(1:3)])
rownames(M.y) <- SiteLookup$siteplot

```


## Unconstrained ordination

This model relates species to two latent variables. No site predictors or random effects are included.

### JAGS model code

```{r}

ordination.code <- "model {
  # Model species occurrence
  for(i in 1:nplots) {
    for(j in 1:nspecies) { 
      eta[i,j] <- inprod(lv.coefs[j,2:(num.lv+1)],lvs[i,])

      Z[i,j] ~ dnorm(lv.coefs[j,1] + eta[i,j], 1)
      y[i,j] ~ dbern(step(Z[i,j]))
    }
  }
  
  ## Latent variables ##
  for(i in 1:nplots) { for(k in 1:num.lv) { lvs[i,k] ~ dnorm(0,1) } } 
  
  ## Process level and priors ##
  for(j in 1:nspecies) { lv.coefs[j,1] ~ dnorm(0,0.1) } ## Separate species intercepts
  
  ## Constraints on latent variables for rotation invariance
  ## (based on scheme used in BORAL)
  for(i in 1:(num.lv-1)) { 
    for(j in (i+2):(num.lv+1)) {
      # Constraints to 0 on upper diagonal
      lv.coefs[i,j] <- 0 
    } 
  }
  
  for(i in 1:num.lv) { 
     # Sign constraints on diagonal elements
    lv.coefs[i,i+1] ~ dnorm(0,0.1)I(0,) 
  }
  
  for(i in 2:num.lv) { 
    for(j in 2:i) { 
      # Free lower diagonals
      lv.coefs[i,j] ~ dnorm(0,0.1) 
    } 
  }
  
  for(i in (num.lv+1):nspecies) { 
    for(j in 2:(num.lv+1)) { 
      # All other elements
      lv.coefs[i,j] ~ dnorm(0,0.1)
    } 
  }
}"

```


### Format data for the model

```{r}

ordination.data <- list(
  y = M.y,
  nsites = max(SiteLookup$site),
  nplots = nrow(M.y),
  nspecies = ncol(M.y),
  num.lv = 2
)

```


### Compile and run the model

```{r}

ModelPath <- here::here("models", "fitted", "unconstrained_ordination.RData")

if (REFIT_MODELS || !file.exists(ModelPath)) {
  modelfit <- R2jags::jags(
    data = ordination.data, 
    inits = list(list(Z = fn_Zinit(M.y))), 
    parameters.to.save = "lvs", 
    model.file = textConnection(ordination.code),
    n.chains = 1, 
    n.iter = 210000, 
    n.burnin = 10000, 
    n.thin = 20
  )
  
  
  samples <- coda::mcmc(modelfit$BUGSoutput$sims.matrix, start = 1, thin = 1)
  save(samples, file = ModelPath)
  
} else {
  samples <- load_from(ModelPath)
}
  
```

Graph the sites and plots in ordination (latent variable) space.


### Ordination graph

```{r}

ii <- str_detect(colnames(samples), "lvs\\[")
x <- apply(as.matrix(samples)[, ii], MARGIN = 2, median)

dat.medians.unconstrained <- as.data.frame(x) %>%
  rename(median = x) %>%
  
  mutate(indices = str_extract(rownames(.), "\\d+\\,\\d")) %>%
  tidyr::separate(indices, into = c("plotindex", "lv"), sep = ",", convert = TRUE) %>%
  mutate(site = 1 + ((plotindex - 1) %/% 3),
         plot = 1 + ((plotindex - 1) %% 3)) %>%
  
  tidyr::pivot_wider(names_from = lv, names_prefix = "lv",
                     values_from = median) %>%
  
  left_join(DAT.floodgroup, by = c("site", "plot"))
  

gg <- ggplot(data = dat.medians.unconstrained, 
             aes(x = lv1, y = lv2, colour = floodgroup)) +
  geom_point(size = 3, alpha = 0.4) +
  geom_text_repel(aes(label = site)) +
  
  scale_colour_discrete("Flood\ngroup") +
  
  lims(x = c(-1, 2), y = c(-2, 1)) +
  coord_equal() +
  
  labs(x = "Latent variable 1", y = "Latent variable 2")

gg_pdf(gg, filename = "unconstrained_ordination.pdf", size = pagesize("A4"))

print(gg)

```


## Latent variables plus flood groups

This model adds the three flood history classification groups.

### JAGS model code

```{r}

floodmodel.code <- "model {
  # Model species occurrence
  for(i in 1:nplots) {
    for(j in 1:nspecies) { 
      eta[i,j] <- inprod(lv.coefs[j,2:(num.lv+1)],lvs[i,]) +
        flood.coefs[j, floodgroup[i]]

      Z[i,j] ~ dnorm(lv.coefs[j,1] + eta[i,j], 1)
      y[i,j] ~ dbern(step(Z[i,j]))
    }
  }
  
  ## Latent variables ##
  for(i in 1:nplots) { for(k in 1:num.lv) { lvs[i,k] ~ dnorm(0,1) } } 
  
  ## Process level and priors ##
  for(j in 1:nspecies) { lv.coefs[j,1] ~ dnorm(0,0.1) } ## Separate species intercepts
  
  ## Constraints on latent variables for rotation invariance
  ## (based on scheme used in BORAL)
  for(i in 1:(num.lv-1)) { 
    for(j in (i+2):(num.lv+1)) {
      # Constraints to 0 on upper diagonal
      lv.coefs[i,j] <- 0 
    } 
  }
  
  for(i in 1:num.lv) { 
     # Sign constraints on diagonal elements
    lv.coefs[i,i+1] ~ dnorm(0,0.1)I(0,) 
  }
  
  for(i in 2:num.lv) { 
    for(j in 2:i) { 
      # Free lower diagonals
      lv.coefs[i,j] ~ dnorm(0,0.1) 
    } 
  }
  
  for(i in (num.lv+1):nspecies) { 
    for(j in 2:(num.lv+1)) { 
      # All other elements
      lv.coefs[i,j] ~ dnorm(0,0.1)
    } 
  }
  
  # species coefficients on flood group
  # drawn from Laplace priors with variance equal to that of 
  # the boral default priors (10).

  laplace.scale <- sqrt(2 / 10)
  for(j in 1:nspecies) { 
    for (kflood in 1:3) {
      flood.coefs[j, kflood] ~ ddexp(0, laplace.scale) 
    }
  }
}"

```


### Format data for the model

```{r}

floodmodel.data <- list(
  y = M.y,
  floodgroup = as.integer(SiteLookup$floodgroup),
  nsites = max(SiteLookup$site),
  nplots = nrow(M.y),
  nspecies = ncol(M.y),
  num.lv = 2
)

```


### Compile and run the model

```{r}

ModelPath <- here::here("models", "fitted", "model_floodgroup.RData")

if (REFIT_MODELS || !file.exists(ModelPath)) {
  modelfit <- R2jags::jags(
    data = floodmodel.data, 
    inits = list(list(Z = fn_Zinit(M.y))), 
    parameters.to.save = c("lvs", "flood.coefs"), 
    model.file = textConnection(floodmodel.code),
    n.chains = 1, 
    n.iter = 210000, 
    n.burnin = 10000, 
    n.thin = 20
  )
  
  
  samples <- coda::mcmc(modelfit$BUGSoutput$sims.matrix, start = 1, thin = 1)
  save(samples, file = ModelPath)
  
} else {
  samples <- load_from(ModelPath)
}
  
```


### Ordination graph

```{r}

ii <- str_detect(colnames(samples), "lvs\\[")
x <- apply(as.matrix(samples)[, ii], MARGIN = 2, median)

dat.medians.flood <- as.data.frame(x) %>%
  rename(median = x) %>%
  
  mutate(indices = str_extract(rownames(.), "\\d+\\,\\d")) %>%
  tidyr::separate(indices, into = c("plotindex", "lv"), sep = ",", convert = TRUE) %>%
  mutate(site = 1 + ((plotindex - 1) %/% 3),
         plot = 1 + ((plotindex - 1) %% 3)) %>%
  
  tidyr::pivot_wider(names_from = lv, names_prefix = "lv",
                     values_from = median) %>%
  
  left_join(DAT.floodgroup, by = c("site", "plot"))
  

gg <- ggplot(data = dat.medians.flood, aes(x = lv1, y = lv2, colour = floodgroup)) +
  geom_point(size = 3, alpha = 0.4) +
  geom_text_repel(aes(label = site)) +
  
  scale_colour_discrete("Flood\ngroup") +
  
  lims(x = c(-1, 2), y = c(-2, 1)) +
  coord_equal() +
  
  labs(x = "Latent variable 1", y = "Latent variable 2")

gg_pdf(gg, filename = "model_floodgroup_ordination.pdf", size = pagesize("A4"))

print(gg)

```

## Latent variables plus stem density

This is the preferred model (see below) with the flood group predictor removed.

### JAGS model code

```{r}

stemsmodel.code <- "model {
  # Model mean stem count by size class within each plot
  # Counts are treated as negative binomial - implemented as
  # Poisson with Gamma-distributed mean
  for (i in 1:length(nstems)) {
    nstems[i] ~ dpois(mu.stems[stemplot[i], sz[i]] * h[i])
    h[i] ~ dgamma(theta[sz[i]], theta[sz[i]])
  }

  # Priors for stem count model
  for (sz in 1:11) {
    # Estimate dispersion independently for each size class
    theta[sz] ~ dexp(1)

    # Overall mean stems (log scale) for each size class
    logmu.size[sz] ~ dnorm(0, 5)

    sd.size[sz] ~ dexp(1)
    tau.size[sz] <- pow(sd.size[sz], -2)
    
    for (i in 1:66) {
      logmu.stems[i, sz] ~ dnorm(logmu.size[sz], tau.size[sz])
      mu.stems[i, sz] <- exp(logmu.stems[i, sz])
    }
  }
  
  # Aggregate modelled stem counts for field size classes into
  # the broader size classes to which species will be related
  for (i in 1:nplots) {
    for (k in 1:nsizeclasses) {
      # MinSize and MaxSize are the start and end indices of field
      # size classes for each aggregated size class
      X.stems[i, k] <- sum(mu.stems[i, (MinSize[k]):(MaxSize[k])])
    }
  }
  
  # Model species occurrence
  for(i in 1:nplots) {
    for(j in 1:nspecies) { 
      eta[i,j] <- inprod(lv.coefs[j,2:(num.lv+1)],lvs[i,]) + 
        siteplot.effect[i] + 
        inprod(stem.coefs[j,], X.stems[i,])
    
      Z[i,j] ~ dnorm(lv.coefs[j,1] + eta[i,j], 1)
      y[i,j] ~ dbern(step(Z[i,j]))

      loglik[(i-1)*(nspecies)+j] <- y[i,j] * log(phi(lv.coefs[j,1] + eta[i,j])) + 
                                 (1 - y[i,j])*log(1 - phi(lv.coefs[j,1] + eta[i,j]))
    }
  }
  
  ## Latent variables ##
  for(i in 1:nplots) { for(k in 1:num.lv) { lvs[i,k] ~ dnorm(0,1) } } 
  
  ## Process level and priors ##
  for(j in 1:nspecies) { lv.coefs[j,1] ~ dnorm(0,0.1) } ## Separate species intercepts
  
  
  ## Nested site-plot random effect
  for(i in 1:nplots) { 
    siteplot.effect[i] ~ dnorm(site.effect[site[i]], pow(siteplot.sigma, -2)) 
  }
  for (i in 1:nsites) {
    site.effect[i] ~ dnorm(0, pow(site.sigma, -2))
  }
  site.sigma ~ dexp(1)
  siteplot.sigma ~ dexp(2) # Weakly constrain site-plot sigma relative to site sigma
  
  
  ## Constraints on latent variables for rotation invariance
  ## (based on scheme used in BORAL)
  for(i in 1:(num.lv-1)) { 
    for(j in (i+2):(num.lv+1)) {
      # Constraints to 0 on upper diagonal
      lv.coefs[i,j] <- 0 
    } 
  }
  
  for(i in 1:num.lv) { 
     # Sign constraints on diagonal elements
    lv.coefs[i,i+1] ~ dnorm(0,0.1)I(0,) 
  }
  
  for(i in 2:num.lv) { 
    for(j in 2:i) { 
      # Free lower diagonals
      lv.coefs[i,j] ~ dnorm(0,0.1) 
    } 
  }
  
  for(i in (num.lv+1):nspecies) { 
    for(j in 2:(num.lv+1)) { 
      # All other elements
      lv.coefs[i,j] ~ dnorm(0,0.1)
    } 
  }
  
  # species coefficients on stem density
  # drawn from Laplace priors with variance equal to that of 
  # the boral default priors (10).

  laplace.scale <- sqrt(2 / 10)
  for(j in 1:nspecies) { 
    for (ksize in 1:nsizeclasses) {
      stem.coefs[j,ksize] ~ ddexp(0, laplace.scale)
    }
  }
}"

```


### Format data for the model

```{r}

# Observed stem densities from pre- and post-thinning surveys. 
# Note - these are the same data as in the file `stems_complete.RData` 
# in the related `red_gum_thinning/stem_densities` project.

path <- here("data", "stems_observed_density.RData")

DAT.stems.observed <- load_from(path) %>%
  # Subset to control and pre-thinning plot data
  dplyr::filter(survey %in% c("control", "pre.thinning")) %>%
  
  arrange(site, plot, subplot) %>%
  
  dplyr::select(site, plot, !!(SIZES$sizeclass)) %>%
  
  tidyr::pivot_longer(-c(site, plot), names_to = "sizeclass", values_to = "nstems") %>%

  dplyr::filter(!is.na(nstems)) %>%
  
  mutate(plotindex = 3*(site-1) + plot,
         sz = match(sizeclass, SIZES$sizeclass)) %>%
  
  dplyr::select(plot = plotindex, sz, nstems)


# Aggregated size classes for modelling

SizeClasses <- c("less10cm", "from10to20cm", "from20to30cm",
                 "from30to40cm", "from40to80cm", "from80cm")

FieldToModelSizeClass <- list(
  1, 2, 3, 4, 5:8, 9:11
)

# JAGS will not work with the lookup list FieldToModelSizeClass
# directly, so we create vector versions
MinSize <- sapply(FieldToModelSizeClass, min)
MaxSize <- sapply(FieldToModelSizeClass, max)

stemsmodel.data <- list(
  y = M.y,
  nsites = max(SiteLookup$site),
  nplots = nrow(M.y),
  nspecies = ncol(M.y),
  
  # site index for each floristic plot
  site = SiteLookup$site,
  
  num.lv = 2,
  
  # stem counts
  stemplot = DAT.stems.observed$plot,
  sz = DAT.stems.observed$sz,
  nstems = DAT.stems.observed$nstems,
  nsizeclasses = length(SizeClasses),
  MinSize = MinSize,
  MaxSize = MaxSize
)

```


### Compile and run the model

```{r}

ModelPath <- here::here("models", "fitted", "model_stems.RData")

if (REFIT_MODELS || !file.exists(ModelPath)) {
  modelfit <- R2jags::jags(
    data = stemsmodel.data, 
    inits = list(list(Z = fn_Zinit(M.y))), 
    parameters.to.save = c("loglik",
                           "lvs", "lv.coefs", 
                           "mu.stems",
                           "siteplot.effect",
                           "siteplot.sigma",
                           "site.sigma",
                           "stem.coefs"), 
    model.file = textConnection(stemsmodel.code),
    n.chains = 1, 
    n.iter = 210000, 
    n.burnin = 10000, 
    n.thin = 20
  )
  
  
  samples <- coda::mcmc(modelfit$BUGSoutput$sims.matrix, start = 1, thin = 1)
  save(samples, file = ModelPath)
  
} else {
  samples <- load_from(ModelPath)
}
  
```


### Ordination graph

```{r}

ii <- str_detect(colnames(samples), "lvs\\[")
x <- apply(as.matrix(samples)[, ii], MARGIN = 2, median)

dat.medians.flood <- as.data.frame(x) %>%
  rename(median = x) %>%
  
  mutate(indices = str_extract(rownames(.), "\\d+\\,\\d")) %>%
  tidyr::separate(indices, into = c("plotindex", "lv"), sep = ",", convert = TRUE) %>%
  mutate(site = 1 + ((plotindex - 1) %/% 3),
         plot = 1 + ((plotindex - 1) %% 3)) %>%
  
  tidyr::pivot_wider(names_from = lv, names_prefix = "lv",
                     values_from = median) %>%
  
  left_join(DAT.floodgroup, by = c("site", "plot"))
  

gg <- ggplot(data = dat.medians.flood, aes(x = lv1, y = lv2, colour = floodgroup)) +
  geom_point(size = 3, alpha = 0.4) +
  geom_text_repel(aes(label = site)) +
  
  scale_colour_discrete("Flood\ngroup") +
  
  lims(x = c(-1, 2), y = c(-2, 1)) +
  coord_equal() +
  
  labs(x = "Latent variable 1", y = "Latent variable 2")

gg_pdf(gg, filename = "model_stems_ordination.pdf", size = pagesize("A4"))

print(gg)

```


## Preferred model

This is the preferred model, with latent variables plus flood groups, stem density and nested random effects (plot within site). See the document `00_preferred_model_stems_species.Rmd` for model code and fitting details. Here we assume that MCMC samples have been saved to file.

```{r}

ModelPath <- here::here("models", "fitted", "model_chosen_mcmc.RData")

if (!file.exists(ModelPath)) 
  stop("Missing file with MCMC samples for chosen model")

samples <- load_from(ModelPath)

# Discard log-likelihood params
ii <- str_detect(colnames(samples), "loglik")
samples <- samples[, !ii]

ii <- str_detect(colnames(samples), "loglik")
post.samples <- as.matrix(samples[, !ii])

rm(samples)
gc()

```


Graph the sites and plots in ordination (latent variable) space.

```{r}

ii <- str_detect(colnames(post.samples), "lvs\\[")
x <- apply(post.samples[, ii], MARGIN = 2, median)

dat.medians.chosen <- as.data.frame(x) %>%
  rename(median = x) %>%
  
  mutate(indices = str_extract(rownames(.), "\\d+\\,\\d")) %>%
  tidyr::separate(indices, into = c("plotindex", "lv"), sep = ",", convert = TRUE) %>%
  mutate(site = 1 + ((plotindex - 1) %/% 3),
         plot = 1 + ((plotindex - 1) %% 3)) %>%
  
  tidyr::pivot_wider(names_from = lv, names_prefix = "lv",
                     values_from = median) %>%
  
  left_join(DAT.floodgroup, by = c("site", "plot"))
  

gg <- ggplot(data = dat.medians.chosen, aes(x = lv1, y = lv2, colour = floodgroup)) +
  geom_point(size = 3, alpha = 0.4) +
  geom_text_repel(aes(label = site)) +
  
  scale_colour_discrete("Flood\ngroup") +
  
  lims(x = c(-1.5, 2), y = c(-1.5, 2)) +
  coord_equal() +
  
  labs(x = "Latent variable 1", y = "Latent variable 2")

gg_pdf(gg, filename = "model_chosen_ordination.pdf", size = pagesize("A4"))

print(gg)

```

## Combined graph with aligned ordinations

```{r}

# Check if package vegan is installed
if (!("vegan" %in% installed.packages()[, "Package"])) {
  stop("package vegan is required for the combined ordination graph")
}


res <- vegan::procrustes(dat.medians.unconstrained[, c("lv1", "lv2")],
                         dat.medians.flood[, c("lv1", "lv2")])

dat.medians.flood.rot <- dat.medians.flood
dat.medians.flood.rot$lv1 <- res$Yrot[,1]
dat.medians.flood.rot$lv2 <- res$Yrot[,2]


res <- vegan::procrustes(dat.medians.unconstrained[, c("lv1", "lv2")],
                         dat.medians.chosen[, c("lv1", "lv2")])

dat.medians.chosen.rot <- dat.medians.chosen
dat.medians.chosen.rot$lv1 <- res$Yrot[,1]
dat.medians.chosen.rot$lv2 <- res$Yrot[,2]


dat.gg <- bind_rows(
  dat.medians.unconstrained %>% mutate(model = "unconstrained"),
  dat.medians.flood.rot %>% mutate(model = "flood"),
  dat.medians.chosen.rot %>% mutate(model = "chosen") ) %>%
  
  mutate(model = factor(model, levels = c("unconstrained", "flood", "chosen")))


gg <- ggplot(data = dat.gg, aes(x = lv1, y = lv2, colour = floodgroup)) +
  geom_point(size = 3, alpha = 0.4) +
  geom_text_repel(aes(label = site)) +
  
  scale_colour_discrete("Flood\ngroup") +
  
  #lims(x = c(-1.5, 2), y = c(-1.5, 2)) +
  coord_equal() +
  
  labs(x = "Latent variable 1", y = "Latent variable 2") +
  
  facet_wrap(~ model)


gg_pdf(gg, filename = "all_ordinations.pdf", size = pagesize("A3", "landscape"))

print(gg)

```



## Variance partitioning

**WARNING** Variance partitioning based on linear predictor values, as calculated below, did not prove useful. The code below is retained in case it is of interest for other projects.

Ubiquitous species which, by definition, were unresponsive to tree stem density, were assigned a large proportion of variance due to stems. This reflected high linear predictor values which were variable but which all corresponded to probabilities > 0.99. Hence the variance was essentially noise. In addition, the use of unscaled predictor values for stem density within size classes was not accounted for in the calculations below.


```{r eval = FALSE}

nplots <- nrow(SiteLookup)
nspp <- nrow(SpeciesSubset)
nsizeclasses <- 6 # aggregated size classes
nfloodgrp <- 3
nlv <- 2
niter <- nrow(post.samples)

# Mapping from field to model aggregate size classes
FieldToModelSizeClass <- list(
  1, 2, 3, 4, 5:8, 9:11
)
names(FieldToModelSizeClass) <- c("less10cm", "from10to20cm", "from20to30cm",
                                  "from30to40cm", "from40to80cm", "from80cm")

get_indices <- function(ptn, n.expected = NULL) {
  ii <- str_detect(colnames(post.samples), ptn)
  if (!is.null(n.expected)) {
    stopifnot(sum(ii) == n.expected)
  }
  ii
}

ii.lvcoefs <- get_indices("lv.coefs", nspp * (1 + nlv))
ii.stemcoefs <- get_indices("stem.coefs", nspp * nsizeclasses)
ii.floodcoefs <- get_indices("flood.coefs", nspp * nfloodgrp)
ii.site.sigma <- get_indices("site.sigma", 1)
ii.siteplot.sigma <- get_indices("siteplot.sigma", 1)

ii.lvs <- get_indices("lvs", nplots * 2)
ii.mu.stems <- get_indices("mu.stems", nplots * 11) # 11 field size classes

m.floodgroup <- matrix(0, nrow = nplots, ncol = nfloodgrp)
ii <- cbind(1:nplots, SiteLookup$floodgroup)
m.floodgroup[ii] <- 1


intercept_var <- lv_var <- stem_var <- 
  flood_var <- raneff_var <- matrix(0, niter, nspp)


for (k in 1:nrow(post.samples)) {
  # species intercepts (col 1) and latent variable coefs (cols 2-3)
  lv.coefs <- matrix(post.samples[k, ii.lvcoefs], nrow = nspp)

  intercept_var[k, ] <- lv.coefs[,1]^2
  lv_var[k, ] <- rowSums(lv.coefs[, 2:(1+nlv)]^2)
  
  # contribution of stem density
  stem.coefs <- matrix(post.samples[k, ii.stemcoefs], nrow = nspp)
  mu.stems <- matrix(post.samples[k, ii.mu.stems], nrow = nplots)
  
  # aggregate modelled field counts to the broader six size classes
  X.stems <- lapply(FieldToModelSizeClass, function(cols) {
    x <- rep(0, nplots)
    for (k in cols) {
      x <- x + mu.stems[,k]
    }
    x
  })
  X.stems <- do.call(cbind, X.stems)
  
  stem.linpred <- tcrossprod(X.stems, stem.coefs)
  stem_var[k, ] <- apply(stem.linpred, 2, var)

  # contribution of flood group
  flood.coefs <- matrix(post.samples[k, ii.floodcoefs], nrow = nspp)
  flood.linpred <- tcrossprod(m.floodgroup, flood.coefs)
  flood_var[k, ] <- apply(flood.linpred, 2, var)

  # contribution of the nested random effect
  raneff_var[k, ] <- post.samples[k, ii.site.sigma]^2 + post.samples[k, ii.siteplot.sigma]^2
}


total.var <- intercept_var + lv_var + stem_var + flood_var + raneff_var


dat.varpart <- SpeciesSubset %>%
  select(workingname, weed, habitat) %>%
  
  mutate(intercept = colMeans(intercept_var / total.var),
         lv = colMeans(lv_var / total.var),
         stem = colMeans(stem_var / total.var),
         flood = colMeans(flood_var / total.var),
         raneff = colMeans(raneff_var / total.var) )

```


For each species, we estimate the relative proportion of variance accounted for by its intercept, latent variables, flood group, stem density and nested random effect.

```{r fig.height=8, eval = FALSE}

dat.gg <- dat.varpart %>%
  tidyr::pivot_longer(cols = c(intercept, lv, stem, flood, raneff),
                      names_to = "varpart",
                      values_to = "value")
  

# arrange names in order of variance accounted for by stem density
name.order <- dat.gg %>%
  filter(varpart == "stem") %>%
  arrange(value) %>%
  mutate(name.habitat = sprintf("%s (%s)", workingname, habitat)) %>%
  select(workingname, name.habitat)
           

dat.gg <- dat.gg %>%
  mutate(label = factor(workingname, 
                        levels = name.order$workingname, 
                        labels = name.order$name.habitat),
         
         varpart = factor(varpart,
                          levels = c("Species intercept" = "intercept", 
                                     "Site-plot effect" = "raneff", 
                                     "Latent variables" = "lv", 
                                     "Flood group" = "flood", 
                                     "Stem density" = "stem")) )

gg <- ggplot(data = dat.gg) +
  geom_bar(aes(x = label, y = value, fill = varpart),
             stat = "identity", width = 0.5, alpha = 0.6) +

    scale_fill_viridis_d(name = "Variance component",
                         begin = 0, end = 0.8, direction = -1) +

    labs(x = "", y = "Proportion of variance") +

    coord_flip() +

    theme(legend.position = "right")


gg_pdf(gg, 
       size = pagesize("A3", "portrait", "cm"),
       filename = "chosen_model_varpart.pdf")

print(gg)

```

