---
title: "Post-hoc GAMS of species responses"
output: 
  word_document: 
    fig_width: 8
    fig_height: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(here)
library(dplyr, warn.conflicts = FALSE)
library(stringr)

library(rjags)
library(R2jags)

library(ggplot2)
library(ggrepel)


# Random number seed
set.seed(42)

# Set this to TRUE to force re-fitting of models if
# cached objects exist. Set to FALSE to use cached outputs.
REFIT_MODELS <- FALSE

# Default theme for graphs
theme_set( theme_bw() + theme(panel.grid = element_blank()))

# Load common functions
source(file = here("scripts", "common_functions.R"))

```


## Data

Lookup tables for study sites and plots, and tree size classes (from the RRGcommon package used for other analyses).

```{r}

SITES <- load_from(here("data", "SITES.RData"))

SIZES <- load_from(here("data", "SIZE_CLASSES.RData"))

```


Response data is a matrix of species occurrences within 9ha experimental plots. The data for each plot are formed by pooling records from the three floristics quadrats located within the plot.

```{r}

DAT.occurrence.long <- load_from( here("data", "site_species_boral.RData") ) %>%
  
  mutate(siteplot = paste0(site, "-", plot9ha)) %>%
  
  distinct(site, plot9ha, siteplot, workingcode)

```


Coding, name and habitat information for species.

```{r}

SPECIES.info <- load_from( here("data", "species_info_boral.RData") )

```


Summary of species prevalence in sites, plots and quadrats.

```{r}

SPECIES.summary <- load_from( here("data", "species_summary_boral.RData") )

```


Flood history plot groupings.

```{r}

DAT.floodgroup <- load_from( here("data", "flood_group_boral.RData") )

```


### Identify a subset of species for modelling

Of the 200 species identified as candidates for modelling, 84 were only recorded in 1 or 2 experimental plots (most of the latter being in a single site). 

The threshold number of occurrences set to include a species for modelling is hard to define, but given the complexity of the models to be fitted, a minimum of 5 plot occurrences is a reasonable starting point.

We also remove Eucalpytus camaldulensis (present in all sites) from the data set.

```{r}

SpeciesSubset <- SPECIES.summary %>%
  dplyr::filter(nplots >= 5) %>%
  
  left_join(SPECIES.info, by = c("workingcode", "workingname")) %>%
  
  # remove red gums
  dplyr::filter(workingname != "Eucalyptus camaldulensis")

cat("Subset with", nrow(SpeciesSubset), "taxa", 
    "of which", sum(SpeciesSubset$weed), "are exotic \n")

```

### Species occurrence data matrix for modelling

```{r}

DAT.occurrence.subset <- DAT.occurrence.long %>%
  dplyr::filter(workingcode %in% SpeciesSubset$workingcode)

# Format response data as a matrix for boral
M.y <- DAT.occurrence.subset %>%
  mutate(present = 1) %>%
  
  tidyr::pivot_wider(names_from = workingcode, 
                     values_from = present, values_fill = 0) %>%

  # Order should be fine but just in case...
  arrange(site, plot9ha)


SiteLookup <- M.y %>%
  dplyr::select(site, plot = plot9ha, siteplot) %>%
  left_join(
    SITES %>% dplyr::select(site, plot, site.quality),
    by = c("site", "plot")
  ) %>%
  
  left_join(DAT.floodgroup, by = c("site", "plot", "siteplot"))


M.y <- as.matrix(M.y[ , -(1:3)])
rownames(M.y) <- SiteLookup$siteplot

```


## Preferred model

### Load model object and get MCMC samples

```{r}

ModelPath <- here::here("models", "fitted", "model_chosen_mcmc.RData")

if (!file.exists(ModelPath)) 
  stop("Missing file with MCMC samples for chosen model")

samples <- load_from(ModelPath)

# Discard log-likelihood params
ii <- str_detect(colnames(samples), "loglik")
samples <- samples[, !ii]

ii <- str_detect(colnames(samples), "loglik")
post.samples <- as.matrix(samples[, !ii])

rm(samples)
gc()

```

### Retrieve posterior samples for mean stem counts

```{r}

# This assumes mu.stems params are ordered by plot (first index) within size 
# class (second index)
# i.e. mu.stems[plot1, sz1], mu.stems[plot2, sz1] ... mu.stems[plot66, sz11]
ii <- str_detect(colnames(post.samples), "mu.stems")

nr <- nrow(post.samples)
dat.mu <- array(NA, dim = c(66, 11, nr))
for (r in 1:nr) dat.mu[,,r] <- post.samples[r,ii]

# Group size classes as per model: 1,2,3,4,5-8,9-11
for (r in 1:nr) {
  dat.mu[,5,r] <- rowSums(dat.mu[,5:8,r])
  dat.mu[,6,r] <- rowSums(dat.mu[,9:11,r])
}
dat.mu <- dat.mu[,1:6,]

# Median posterior stem densities for actual sites
dat.actual <- apply(dat.mu, MARGIN = 1:2, median) %>%
  as.data.frame()

colnames(dat.actual) <- c("<10cm", "<20cm", "<30cm", "<40cm", "<80cm", ">80cm")

```


## Individual species GAMS

### Plotting function

The function takes data for a single species and fits a GAM smooth for each stem size class. 

It also fits a trend line (shown in red on the graphs) derived from the posterior distribution of the preferred model. This is done for each species by:

  1. Get the MCMC samples for the species intercept and the stem density coefficients.
  
  2. Calculate a linear predictor value based on the mean stem density values for each size class individually: lp = b0 + b1[sizeclass] * mean.density
  
  3. Calculate median value of the linear predictor for each size class and transform to a probability using probit link.
  
The intention is to show the marginal effect of each stem size class on probability of occurrence.


```{r}

fn_plot <- function(spindex) {
  y <- M.y[, spindex]
  
  # species intercepts
  ptn <- sprintf("lv.coefs\\[%d,1\\]", spindex)
  ii <- str_detect(colnames(post.samples), ptn)
  b0 <- post.samples[, ii]
  
  # species coefficients on stem density
  ptn <- sprintf("stem.coefs\\[%d,", spindex)
  ii <- str_detect(colnames(post.samples), ptn)
  b1 <- post.samples[, ii]
  
  X <- as.matrix(dat.actual)
  
  # GLM prediction based only on intercept and stem density
  pfit <- lapply(1:ncol(dat.actual), function(k) {
    lp <- b0 + tcrossprod(b1[,k], X[,k])
    lp <- apply(lp, MARGIN = 2, quantile, probs = c(0.25, 0.5, 0.75))
    lp <- t(lp)
    colnames(lp) <- c("lwr", "mid", "upr")
    
    out <- data.frame(sizeclass = colnames(X)[k], 
                      plotindex = 1:nrow(X))
    
    cbind(out, pnorm(lp))
  })
  
  pfit <- do.call(rbind, pfit)
  
  dat.gg <- cbind(dat.actual, plotindex = 1:nrow(M.y), y = M.y[, spindex]) %>%
    tidyr::pivot_longer(-c(plotindex, y), names_to = "sizeclass") %>%
    mutate(present = (y == 1)) %>%
    left_join(pfit, by = c("plotindex", "sizeclass"))
  
  ggplot(data = dat.gg, aes(x = value, y = y)) +
    geom_point(aes(shape = present), colour = "grey60", show.legend = FALSE) +
    
    geom_smooth(method = "gam", formula = y ~ s(x, k=4, bs="ts"), 
                method.args = list(family = binomial(), method = "REML", select=TRUE),
                alpha = 0.4) +
    
    geom_smooth(method = "glm", formula = y ~ x,
                method.args = list(family = binomial()),
                se = FALSE, linetype = "dashed") +
    
    geom_ribbon(aes(ymin = lwr, ymax = upr), 
                fill = "red", alpha = 0.2) +
    
    geom_line(aes(y = mid), colour = "red") +
    
    scale_y_continuous(limits = c(0,1)) +
    scale_shape_manual(values = c(1, 16), drop=FALSE) +
    
    labs(title = SpeciesSubset$workingname[spindex]) +
    
    facet_wrap(~ sizeclass, scales = "free_x")
}

```


### Graph each species

```{r echo=TRUE}

Nspp <- nrow(SpeciesSubset)

for (i in seq(1, Nspp, 4)) {
  ggs <- lapply(i:min(i+3, Nspp), fn_plot)
  
  gg <- patchwork::wrap_plots(ggs, ncol = 2)
  
  print(gg)
}

```



